---
title: "STAT 684 Project - Multivariate Time Series Analysis of U.S. Beef and Chicken Prices"
author: "Ken Marciel"
date: "4/26/2022"
output: html_document
---

## Packages

```{r packages, message=FALSE, warning=FALSE}
library(dplyr)
library(dygraphs)
library(forecast)
library(h2o)
library(imputeTS)
library(lubridate)
library(plotly)
library(readxl)
library(TSstudio)
```


## Univariate Time Series Data for U.S. Beef Prices

Data set: U.S. average price of 100% ground beef in dollars per pound
Period: January 1, 1984, to March 1, 2022
Seasonally adjusted: No
Source: Economic Research data from the Federal Reserve Bank of St. Louis (FRED)
Series: APU0000703112
Web address: https://fred.stlouisfed.org/series/APU0000703112

```{r beefDataSet, include=FALSE}
## Obtain data
# set working directory to location of downloaded data files
path <- 
"C:/Users/keoka/OneDrive - Texas A&M University/Courses/STAT_684/Project/Data"
setwd(path)
bf_df <- read_excel("APU0000703112.xls") # data series for beef prices

## Examine data
class(bf_df) # table, data frame
dim(bf_df) # 469 rows, 2 columns
str(bf_df); summary(bf_df) # tibble with two character variables
head(bf_df,10+5); tail(bf_df,5) # view first 5 and last 5 observations
names(bf_df) # variable names: "FRED Graph Observations" and "...2"

## Reformat data
bf_df <- bf_df[11:469,] # exclude description of data set in first 10 rows
names(bf_df) <- c("date","beef price") # rename variables
bf_df[,1] <- as.Date(as.numeric(unlist(bf_df[,1])), # character to numeric date
                  origin = as.Date("1899-12-30")) # align to Excel origin point
bf_df[,2] <- round(as.numeric(unlist(bf_df[,2])),3) # character to numeric values
str(bf_df); summary(bf_df) # tibble with date variable and numeric variable
which(is.na(bf_df[,2])) # no missing values
which(bf_df[,2]==0) # missing value was converted to zero value in row 346
bf_df[341:350,] # view zero value in row 346
bf_df[346,2] <- NA # convert zero value back to missing (NA) value
bf_df[341:350,] # view missing value (NA) in row 346
# impute missing value using linear interpolation
bf_df[346,2] <- na_interpolation(ts(bf_df))[346,2]
bf_df[341:350,] # view imputed value in row 346

## Convert data frome to univariate time series object with defined attributes
bf_ts <- ts(data = bf_df[,2], # series values
         start = c(1984,1), # time of first observation: January 1, 1984
         end = c(2022,3), # time of last observation: March 1, 2022
         frequency = 12) # series frequency
ts_info(bf_ts) # ts object with 1 variable and 459 observations
```


## Univariate time series data for U.S. chicken prices

Data set: U.S. average price of fresh, whole chicken in dollars per pound
Period: January 1, 1980, to March 1, 2022
Seasonally adjusted: No
Source: Economic Research data from the Federal Reserve Bank of St. Louis (FRED)
Series: APU0000706111
Web address: https://fred.stlouisfed.org/series/APU0000706111

```{r chickenDataSet, include=FALSE}
## Obtain data
# set working directory to location of downloaded data files
path = 
"C:/Users/keoka/OneDrive - Texas A&M University/Courses/STAT_684/Project/Data"
setwd(path)
ch_df = read_excel("APU0000706111.xls") # data series for chicken prices

## Examine data
class(ch_df) # table, data frame
dim(ch_df) # 517 rows, 2 columns
str(ch_df); summary(ch_df) # tibble with two character variables
head(ch_df,10+5); tail(ch_df,5) # view first 5 and last 5 observations
names(ch_df) # variable names: "FRED Graph Observations" and "...2"

## Reformat data
ch_df <- ch_df[11:517,] # exclude description of data set in first 10 rows
names(ch_df) <- c("date","chicken price") # rename variables
ch_df[,1] <- as.Date(as.numeric(unlist(ch_df[,1])), # character to numeric date
                  origin = as.Date("1899-12-30")) # align to Excel origin point
ch_df[,2] <- round(as.numeric(unlist(ch_df[,2])),3) # character to numeric values
str(ch_df); summary(ch_df) # tibble with date variable and numeric variable
which(is.na(ch_df[,2])) # no missing values
which(ch_df[,2]==0) # missing value was converted to zero value in row 485
ch_df[481:490,] # view zero value in row 485
ch_df[485,2] <- NA # convert zero value back to missing (NA) value
ch_df[481:490,] # view missing value (NA) in row 485
# impute missing value using linear interpolation
ch_df[485,2] <- na_interpolation(ts(ch_df))[485,2]
ch_df[481:490,] # view imputed value in row 485

## Convert data frome to univariate time series object with defined attributes
ch_ts <- ts(data = ch_df[,2], # series values
         start = c(1980,1), # time of first observation: January 1, 1980
         end = c(2022,3), # time of last observation: March 1, 2022
         frequency = 12) # series frequency
ts_info(ch_ts) # ts object with 1 variable and 507 observations
```


## Multivariate time series data for U.S. beef and chicken prices

```{r multivariateDataSet, include=FALSE}
## Create multivariate time series object
bfch_ts <- ts(data = cbind(bf_df$`beef price`, # series values
                           ch_df$`chicken price`[49:507]),
              # minimum common date
              start = c(year(min(bf_df$date)), month(min(bf_df$date))),
              frequency = 12) # series frequency
ts_info(bfch_ts) # mts object with 2 variable and 459 observations
```


## Exploratory data analysis

```{r explore, echo=FALSE}
## Plot time series with "plot" function from R built-in "graphics" package
plot.ts(bfch_ts, # multivariate time series object
        plot.type = "multiple",
        main = "U.S. Average Price of Beef vs. Chicken",
        ylab = c("Beef","Chicken"),
        xlab = "Years")
plot.ts(bfch_ts, # multivariate time series object
        plot.type = "single",
        main = "U.S. Average Price of Beef vs. Chicken",
        ylab = c("Beef","Chicken"),
        xlab = "Years")

## Plot time series with "ts_plot" function from "TSstudio" package
ts_plot(bfch_ts, # multivariate time series object
        type = "multiple",
        title = "U.S. Average Price of Beef vs. Chicken",
        Ytitle = "Cost per Pound",
        Xtitle = "Years",
        Xgrid = TRUE,
        Ygrid = TRUE)
ts_plot(bfch_ts, # multivariate time series object
        type = "single",
        title = "U.S. Average Price of Beef vs. Chicken",
        Ytitle = "Cost per Pound",
        Xtitle = "Years",
        Xgrid = TRUE,
        Ygrid = TRUE)

## Plot time series with "dygraph" function from "dygraphs" package
dygraph(bfch_ts, # multivariate time series object
        main = "U.S. Average Price of Beef vs. Chicken") %>%
  dyAxis("y",label = "Cost per Pound") %>%
  dySeries("Series 1",axis='y',label = "Ground Beef", color="red") %>%
  dySeries("Series 2",axis = 'y',label = "Whole Chicken",color = "orange") %>%
  dyLegend()
```

