---
title: "STAT 684 - Statistical Consulting"
author: "Ken Marciel"
date: "4/26/2022"
output: html_document
---

## Packages

```{r packages, message=FALSE, warning=FALSE}
library(dplyr)
library(dygraphs)
library(forecast)
library(h2o)
library(imputeTS)
library(lubridate)
library(plotly)
library(readxl)
library(TSstudio)
```


## Univariate Time Series Data for U.S. Beef Prices

Data set: U.S. average price of 100% ground beef in dollars per pound
Period: January 1, 1984, to March 1, 2022
Seasonally adjusted: No
Source: Economic Research data from the Federal Reserve Bank of St. Louis (FRED)
Series: APU0000703112
Web address: https://fred.stlouisfed.org/series/APU0000703112

```{r dataSet, include=FALSE}
## Obtain data
# set working directory to location of downloaded data files
path <- 
"C:/Users/keoka/OneDrive - Texas A&M University/Courses/STAT_684/Project/Data"
setwd(path)
bf_df <- read_excel("APU0000703112.xls") # data series for beef prices

## Examine data
class(bf_df) # table, data frame
dim(bf_df) # 469 rows, 2 columns
str(bf_df); summary(bf_df) # tibble with two character variables
head(bf_df,10+5); tail(bf_df,5) # view first 5 and last 5 observations
names(bf_df) # variable names: "FRED Graph Observations" and "...2"

## Reformat data
bf_df <- bf_df[11:469,] # exclude description of data set in first 10 rows
names(bf_df) <- c("date","beef price") # rename variables
bf_df[,1] <- as.Date(as.numeric(unlist(bf_df[,1])), # character to numeric date
                  origin = as.Date("1899-12-30")) # align to Excel origin point
bf_df[,2] <- round(as.numeric(unlist(bf_df[,2])),3) # character to numeric values
str(bf_df); summary(bf_df) # tibble with date variable and numeric variable
which(is.na(bf_df[,2])) # no missing values
which(bf_df[,2]==0) # missing value was converted to zero value in row 346
bf_df[341:350,] # view zero value in row 346
bf_df[346,2] <- NA # convert zero value back to missing (NA) value
bf_df[341:350,] # view missing value (NA) in row 346
# impute missing value using linear interpolation
bf_df[346,2] <- na_interpolation(ts(bf_df))[346,2]
bf_df[341:350,] # view imputed value in row 346

## Convert data frome to univariate time series object with defined attributes
bf_ts <- ts(data = bf_df[,2], # series values
         start = c(1984,1), # time of first observation: January 1, 1984
         end = c(2022,3), # time of last observation: March 1, 2022
         frequency = 12) # series frequency
ts_info(bf_ts) # ts object with 1 variable and 459 observations
```

This univariate time series data set contains the average monthly values for the price of 100% ground beef in the United States, measured in dollars per pound, from January 1st, 1984, to March 1st, 2022, for a total of 459 months. The missing value for October 1st, 2012, was imputed using linear interpolation.


## Exploratory Data Analysis

### Plot of observed values

```{r explore, echo=FALSE}
## Plot time series with "plot" function from R built-in "graphics" package
plot.ts(bf_ts, # univariate time series object
        main="U.S. Average Price of 100% Ground Beef",
        ylab="Cost per Pound",
        xlab="Years")

## Plot time series with "ts_plot" function from "TSstudio" package
ts_plot(bf_ts, # univariate time series object
        title = "U.S. Average Price of 100% Ground Beef",
        Ytitle = "Cost per Pound",
        Xtitle = "Years",
        Xgrid = TRUE,
        Ygrid = TRUE)

## Plot time series with "dygraph" function from "dygraphs" package
dygraph(bf_ts,
        main = "U.S. Average Price of 100% Ground Beef",
        ylab = "Cost per Pound",
        xlab = "Years")
```


### Decomposition of time series data

```{r decomposition, echo=FALSE}
## Classical decomposition of time series
bf_dc <- decompose(bf_ts) # function from R built-in "stats" package
ts_decompose(bf_ts) # function from "TSstudio" package
class(bf_dc); summary(bf_dc) # review attributes of decomposed time series
plot(bf_dc) # plot decomposed time series
```

The time series has a growing trend with an embedded cycle, which are both apparent in the observed series. The most recent cycle started just before 2010, near the end of the Great Recession that began in 2008. The seasonal component is not apparent in the observed series. The impact of the COVID-19 pandemic from 2020 to 2022 is conspicuous in both the observed series and the random component. The time series plot can be decomposed to show the trend (including cycle), seasonal, and random components.


### Seasonality analysis

```{r seasonality, echo=FALSE, warning=FALSE}
## Time series heatmap
ts_heatmap(bf_ts, title = "Heatmap - US Average Price of Ground Beef")

## Seasonality plots
ggseasonplot(bf_ts) # seasonal plot
ggseasonplot(bf_ts,polar = TRUE) # polar plot
ts_seasonal(bf_ts,type = "normal")
ts_seasonal(bf_ts,type = "cycle")
ts_seasonal(bf_ts,type = "box")
ts_seasonal(bf_ts,type = "all")

## Box plot of the seasonal component of the detrended series
bf_dt <- bf_ts - decompose(bf_ts)$trend # remove trend
ts_seasonal(bf_dt,type = "box") # box plot
```

The heatmap shows evidence of cyclic behavior (across the vertical bars), but not seasonal behavior (across the horizontal bars). All four seasonality plots lack evidence for seasonal behavior in the time series based on the following behavior:

- Horizontal lines in the standard plot
- Rope appearance in the cycle plot
- Horizontal pattern across the box plots
- Circular spiral pattern in the polar plot


### Correlation analysis

```{r correlation, echo=FALSE, warning=FALSE}
par(mfrow=c(1,2)) # display two plots in one row
acf(bf_ts, lag.max = 60) # autocorrelation function
pacf(bf_ts, lag.max = 60) # partial autocorrelation function
ts_cor(bf_ts) # acf and pacf
ts_lags(bf_ts) # lag plots
ts_lags(bf_ts,lags = c(12,24,36,48)) # seasonal lags
checkresiduals(bf_ts) # residual analysis
Box.test(bf_ts,type = "Ljung") # Ljung-Box test of autocorrelation
```

The correlation of the series with its lags is decaying gradually over time, with no apparent seasonal component.

The lack of seasonality makes sense given that beef is a food eaten year-round in the United States.


## Forecasting Strategies

### Linear Regression

The time series data set was partitioned into a training set consisting of the values of the first 447 months, and a test set consisting of the last 12 months.

```{r partition, echo=FALSE}
## Linear regression forecasting model
h <- 12 # forecast horizon = last 12 observations
bf_split <- ts_split(bf_ts,sample.out = h) # partition of data set
bf_train <- bf_split$train # training set
bf_test <- bf_split$test # test set
bf_lr1 <- tslm(bf_train ~ season + trend + I(trend^2)) # trained model
summary(bf_lr1)

## Forecast the next 12 months (testing set)
fc_lr1 <- forecast(bf_lr1,h=12)
plot_forecast(fc_lr1)

## Evaluate model performance
accuracy(fc_lr1,bf_test)
test_forecast(actual = bf_ts,
              forecast.obj = fc_lr1,
              test = bf_test)

## Analyze model residuals
checkresiduals(fc_lr1)
```

The model coefficients for the intercept and the trend components are statistically significant at a level of less than 0.001. None of the coefficients for the seasonal components are statistically significant.

The MAPE is 8.35% on the training set and 3.07% on the test set. This imbalance suggests that the predictors may be overfitting the model, consistent with the high adjusted R-squared value of 0.93. Alternatively, the imbalance may be due to the dates of the test set occurring during the global pandemic, which is an event that is unprecedented in the time period of the data set from 1984 to the present, resulting in outliers.

The Ljung-Box test has a statistically-significant p-value, rejecting the null hypothesis that the random component is white noise. This is confirmed by visual analysis of the residuals, which show significant correlation in the model between the series and its lags. This means that the model does not capture a majority of the variation patterns of the series. Therefore, it is not a valid model for consideration. However, we will use its MAPE score of 3.07% as a benchmark to evaluate the performance of the other models that we will train.


### Exponential Smoothing Models

Holt-Winters model

The time series data set was partitioned into a training set consisting of the values of the first 447 months, and a test set consisting of the last 12 months.

```{r holtWinters, echo=FALSE, warning=FALSE}
## Create training and testing partitions
bf_split <- ts_split(bf_ts,12)
bf_train <- bf_split$train
bf_test <- bf_split$test

## Forecast the last 12 months of the series (testing set)
bf_hw1 <- HoltWinters(bf_train) # train a model
bf_hw1 # review the parameters and error rate of the trained model

## Forecast the next 12 months (testing set)
fc_hw1 <- forecast(bf_hw1,h=12)
plot_forecast(fc_hw1)

## Evaluate model performance
accuracy(fc_hw1,bf_test)
test_forecast(actual = bf_ts,
              forecast.obj = fc_hw1,
              test = bf_test)

## Analyze model residuals
checkresiduals(fc_hw1)
```

The Holt-Winters model is mainly learning from the level and seasonal update (with $\alpha=0.94$ and $\gamma=1$). On the other hand, there is essentially no learning from the trend value ($\beta=0.008$). The accuracy metrics of the model are imbalanced, with an MAPE of 1.9% in the training set and 9.32% in the testing set. This makes sense given the global pandemic and overlapping period of acute inflation occurring from early 2020 to the present, which have no precedent since the beginning of the training set in 1984. This can be seen in the plot of model performance. While the plot shows a good fit to the peak in 2020, most of the error comes from underestimating the following peak in 2021.

Residual analysis suggests that the residuals are white noise, so we can conclude that this is a valid forecasting model.

We will next train the Holt-Winters model using a grid search. We will start with a shallow search with larger increments for the tuning parameters. This will narrow the search area for a deeper search of the tuning parameters. The shallow search will consist of backtesting on the training data set, with an expanding window of six different periods spaced six months apart. Each of the three tuning parameters will be initialized to a range of 0 to 1, with an increment of 0.1.

```{r hwShallowGrid, echo=FALSE, warning=FALSE}
shallow_grid <- ts_grid(bf_train,
                        model = "HoltWinters",
                        periods = 6, # number of backtesting periods
                        window_length = NULL, # expanding window
                        window_space = 6, # length between training partitions
                        window_test = 12, # length of testing partition
                        # tuning parameters
                        hyper_params = list(alpha = seq(0,1,0.1),
                                            beta = seq(0,1,0.1),
                                            gamma = seq(0,1,0.1)))
shallow_grid$grid_df[1:10,] # sorted by lowest to highest mean error rate
```

The table displays the results of the shallow grid search. The models are sorted from best to worst, according to the combination of tuning parameters having the lowest mean error rates. The optimal range of $\alpha$ varies between 0.2 and 0.3, $\beta$ is constant at 0.2, and $\gamma$ is between 0.1 and 0.7. This will help us narrow the ranges of parameter values for a deeper grid search.

```{r hwDeepGrid, echo=FALSE}
deep_grid <- ts_grid(bf_train,
                     model = "HoltWinters",
                     periods = 6, # number of backtesting periods
                     window_length = NULL, # expanding window
                     window_space = 6, # length between training partitions
                     window_test = 12, # length of testing partition
                     # tuning parameters
                     hyper_params = list(alpha = seq(0.2,0.3,0.01),
                                         beta = seq(0.15,0.25,0.01),
                                         gamma = seq(0.1,0.7,0.01)))
deep_grid$grid_df[1:10,] # sorted from lowest to highest mean error rate
```

The error range of the top 10 models has dropped slightly compared to the shallow search. The next step is to retrain the HW model using the optimal values of the smoothing parameters from the deep grid search.

```{r hwRetrain, echo=FALSE, warning=FALSE}
## Retrain HW model using optimal parameter values from deep grid search
bf_hw2 <- HoltWinters(bf_train, # training set
                          # tuning parameters from deep grid search
                          alpha = deep_grid$alpha,
                          beta = deep_grid$beta,
                          gamma =  deep_grid$gamma)
bf_hw2 # review the parameters and error rate of the retrained model

## Forecast the next 12 months (testing set)
fc_hw2 <- forecast(bf_hw2,h=12)
plot_forecast(fc_hw2)

## Evaluate the model's performance with testing set
accuracy(fc_hw2,bf_test)
test_forecast(actual = bf_ts,
              forecast.obj = fc_hw2,
              test = bf_test)

## Analyze model residuals
checkresiduals(fc_hw2)
```

As you can see from the plot of fitted and forecasted values, the HW model obtained from the grid search underestimated the 2021 peak, with an MAPE score of 10.5% on the test set. Correlation analysis suggests that the random component is not white noise, meaning that this is not a valid model for consideration.


## Forecasting with ARIMA Models

### Transforming a non-stationary series into a stationary series

```{r stationarity, echo=FALSE}
## Plot of raw time series
ts_plot(bf_ts,
        title = "Monthly Beef Prices 1984-2022",
        Ytitle = "Cost per Pound",
        Xtitle = "Year")

## Log transformation of time series
ts_plot(log(bf_ts),
        title = "Beef Price Series - Log Transformation",
        Ytitle = "Log of Cost per Pound",
        Xtitle = "Year")

## First order differencing
ts_plot(diff(bf_ts,lag=1),
        title = "Beef Price Series - First Differencing",
        Ytitle = "Differencing of Cost per Pound",
        Xtitle = "Year")

## First order and seasonal differencing
ts_plot(diff(diff(bf_ts,lag=1),12),
        title = "Beef Price Series - First and Seasonal Differencing",
        Ytitle = "Differencing of Cost per Pound",
        Xtitle = "Year")

## Log transformation and first order differencing
ts_plot(diff(log(bf_ts),lag=1),
        title = "Beef Price Series - First Differencing with Log Transformation",
        Ytitle = "Differencing/Log of Cost per Pound",
        Xtitle = "Year")

## Log transformation with first order and seasonal differencing
ts_plot(diff(diff(log(bf_ts),lag=1),12),
        title = 
          "Beef Price Series - First and Seasonal Differencing with Log Transformation",
        Ytitle = "Differencing/Log of Cost per Pound",
        Xtitle = "Year")
```

The log transformation with first-order differencing did the best job of transforming the series to a stationary state and stabilizing the series variation. We will next use this sequence of transformations to manually fit an ARIMA model.


### Fitting an ARIMA model with a manual process

Autoregressive Integrated Moving Average (ARIMA)

```{r manualARIMA, echo=FALSE}
par(mfrow=c(1,2)) # display two plots in one row
acf(diff(log(bf_ts))) # autocorrelation function (ACF)
pacf(diff(log(bf_ts))) # partial autocorrelation function (PACF)
ts_cor(diff(log(bf_ts))) # acf and pacf

## Train an ARIMA(1,1,0) model
arima_m1 <- arima(log(bf_ts),order = c(1,1,0))
summary(arima_m1)
checkresiduals(arima_m1)
arima_m1_fc <- forecast(arima_m1,h=12)
plot_forecast(arima_m1_fc)

## Train an ARIMA(0,1,1) model
arima_m2 <- arima(log(bf_ts),order = c(0,1,1))
summary(arima_m2)
checkresiduals(arima_m2)
arima_m2_fc <- forecast(arima_m2,h=12)
plot_forecast(arima_m2_fc)

## Train an ARIMA(1,1,1) model
arima_m3 <- arima(log(bf_ts),order = c(1,1,1))
summary(arima_m3)
checkresiduals(arima_m3)
arima_m3_fc <- forecast(arima_m3,h=12)
plot_forecast(arima_m3_fc)
```

A log transformation was applied to the series, followed by first-order differencing. The transformed series cuts off after the first lag in the plots of both the autocorrelation function (ACF) and the partial autocorrelation function (PACF). There is no apparent seasonality pattern. The lags do not appear to tail off in either plot. Therefore, the corresponding models that were trained are ARIMA(1,1,0), ARIMA(0,1,1), and ARIMA(1,1,1).

All three models have nearly identical performance in terms of error metrics and behavior of the residuals. The ARIMA(1,1,1) had a MAPE score of 3.31%, being the lowest of the three models by one thousandth of one percent. Furthermore, the ARIMA(1,1,1) had the most statistically significant coefficients of the three models. 
The plot of residuals from the ARIMA(1,1,1) model show a pattern of random oscillation around zero with stable variation. There are a few outliers, with the most obvious one corresponding to the global pandemic from 2020 to 2022. Otherwise, there is apparent nonrandom pattern in the residuals. The ACF plot confirms the lack of significant autocorrelation with lagged values. The density plot shows that the errors are normally distributed. The Ljung-Box test failed to reject the null hypothesis that the lags are not correlated. Therefore, we conclude that the random component of the model is white noise, meaning that ARIMA(1,1,1) is a valid forecasting model for this data series.


### Fitting an ARIMA model with an automated tuning process

```{r autoARIMA, echo=FALSE}
## Automated ARIMA model tuning using default arguments
arima_a1 <- auto.arima(bf_train) # tune model using training set
summary(arima_a1)
checkresiduals(arima_a1)
arima_a1_fc <- forecast(arima_a1,h=12)
plot_forecast(arima_a1_fc)

## Automated ARIMA model tuning using robust search
arima_a2 <- auto.arima(bf_train, # training set
                       # limit order of model to six
                       max.order = 5, # p+q+P+Q=5
                       d=1, # non-seasonal differencing
                       D=0, # seasonal differencing
                       stepwise = FALSE, # search all possible combinations
                       approximation = FALSE) # for more accurate calculations
summary(arima_a2)
checkresiduals(arima_a2)

## Forecast the next 12 months (testing set)
arima_a2_fc <- forecast(arima_a2,h=12)
plot_forecast(arima_a2_fc)
```

With its default parameters, the automated tuning process fitted an ARIMA(0,1,0) model that includes a drift term. With a robust search, the automated tuning process fit a ARIMA(0,1,0)(0,0,2)[12] with drift. Performance of both models are nearly identical in terms of error metrics and residual behavior. Therefore, ARIMA(0,1,0) with drift is the parsimonious model, having an MAPE score of 1.78%, and residual analysis indicating that the random component is white noise.


### Linear regression with ARIMA errors

We will next train a linear regression model having the following three predictors: trend, 12-month seasonal lag, and a categorical variable for month of the year. Additionally, the errors will be modeled using the ARIMA procedure.

```{r errorARIMA, echo=FALSE}
## Prepare data and create new features for the series
bf_df$lag12 <- dplyr::lag(bf_df$`beef price`,n=12) # seasonal lag
# seasonal component
bf_df$month <- factor(month(bf_df$date,label=TRUE),ordered = FALSE)
# marginal change in series from moving in time by one month
bf_df$trend <- 1:nrow(bf_df)

## Split the series into training and testing partitions
train_df <- bf_df[1:(nrow(bf_df)-12),]
test_df <- bf_df[(nrow(bf_df)-12+1):nrow(bf_df),]

## Train linear regression model with ARIMA errors using xreg argument
arima_e1 <- auto.arima(bf_train, # training set
                       # change month from categorical to binary variable
                       # drop first column (category)
                       xreg = cbind(model.matrix(~month,train_df)[,-1],
                                    train_df$trend,
                                    train_df$lag12),
                       seasonal = TRUE, # search seasonal and non-seasonal models
                       stepwise = FALSE, # search all possible combinations
                       approximation = FALSE) # for more accurate calculations
summary(arima_e1)
checkresiduals(arima_e1)

## Forecast the next 12 months (testing set)
arima_e1_fc <- forecast(arima_e1,xreg = cbind(model.matrix(~month,test_df)[,-1],
                                              test_df$trend,
                                              test_df$lag12))
plot_forecast(arima_e1_fc)
```

The linear regression model with ARIMA errors has an MAPE score of 1.78%, and the correlation analysis indicates that the random component is white noise.


## Forecasting with Machine Learning Models

We will next use several machine learning regression models. Given that the purpose is to obtain a short-term forecast of 12 months, using the entire series may add noise to the model from previous cycles. Therefore, we will instead subset the series to only include the most recent cycle, beginning in January 2010, following the end of the 2008 economic crisis.

```{r subsetData, echo=FALSE}
## Subset time series object and transform to a data frame
bf_df_st <- ts_to_prophet(window(bf_ts,start=c(2010,1)))

## Examine data subset
class(bf_df_st); dim(bf_df_st) # data frame with 147 rows, 2 columns
str(bf_df_st); summary(bf_df_st) # date variable and numeric variable
head(bf_df_st,5); tail(bf_df_st,5) # view first 5 and last 5 observations
names(bf_df_st) # variable names: "ds" and "y"

## Reformat data
names(bf_df_st) <- c("date","y") # rename variables
head(bf_df_st,5) # view first five observations

## Plot time series
ts_plot(bf_df_st, # data frame
        title = "US Average Beef Price Since January 2010",
        Ytitle = "Cost per Pound",
        Xtitle = "Year")
```


### Feature engineering

We will next create new features to be used as informative input for the model.

```{r featureEngineering, echo=FALSE}
## Create new features and add to time series data frame 
bf_df_st <- bf_df_st %>% 
  # categorical variable for month of year to capture seasonality
  mutate(month=factor(month(date,label = TRUE),ordered = FALSE),
         # variable for seasonal lag of 12 months
         lag12=lag(y,n=12)) %>% filter(!is.na(lag12))
bf_df_st$trend <- 1:nrow(bf_df_st) # trend component
bf_df_st$trend_sqr <- bf_df_st$trend^2 # second-degree polynomial of trend
str(bf_df_st); head(bf_df_st) # view structure of data frame with new features
```


### Training, testing, and model evaluation

To allow for model comparison, follow the same procedure used for the previous models to create the training and testing partitions. It will also be necessary to create inputs for the forecast itself.

```{r mlPartition, echo=FALSE}
## Create training and testing partitions of the data subset
h <- 12 # forecast horizon = 12 months
train_df_st <- bf_df_st[1:(nrow(bf_df_st)-h),] # exclude last 12 months
test_df_st <- bf_df_st[(nrow(bf_df_st)-h+1):nrow(bf_df_st),] # last 12 months

## Create data frame with the dates of the following 12 months
fc_df_st <- data.frame(
  date=seq.Date(from = max(bf_df_st$date)+months(1),length.out = h,by="month"),
  # trend component
  trend=seq(from=max(bf_df_st$trend)+1,length.out=h,by=1))

## Build the rest of the features
fc_df_st$trend_sqr <- fc_df_st$trend^2 # second-degree polynomial of trend
# categorical variable for month of year to capture seasonality
fc_df_st$month <- factor(month(fc_df_st$date,label = TRUE),ordered = FALSE)
# extract last 12 observations and assign as future lags of the series
fc_df_st$lag12 <- tail(bf_df_st$y,12)
str(fc_df_st); head(fc_df_st) # view structure of data frame with new features
```


### Model benchmark

We will use a linear regression model as a benchmark for the machine learning models.

```{r benchmark, echo=FALSE}
## Train the linear regression model on the training partition
bf_lr2 <- lm(y ~ month + lag12 + trend + trend_sqr, data = train_df_st)
summary(bf_lr2) # review the model details

## Predict the corresponding values of the series on the testing partition
test_df_st$yhat <- predict(bf_lr2,newdata = test_df_st)

## Evaluate the performance of the model on the testing partition
mape_lr2 <- mean(abs(test_df_st$y-test_df_st$yhat)/test_df_st$y)
mape_lr2

## Analyze model residuals
checkresiduals(bf_lr2)
```

The residual plot shows a nonrandom pattern. The correlation plot shows that the series is dependent on its lags. The density plot shows a right-skewed distribution. Therefore, it is not a valid forecasting model. However, we will use its MAPE score of 10.04% as a benchmark for the performance of the machine learning models.


### Starting an h2o cluster

```{r h2oCluster, echo=FALSE, message=FALSE, warning=FALSE}
## Set the in-memory cluster with the H2O function
h2o.init() # provides information about the cluster's setup

## Transform data frame objects to h2o clusters
train_h <- as.h2o(train_df_st) # training set
test_h <- as.h2o(test_df_st) # test set
forecast_h <- as.h2o(fc_df_st) # future values of series inputs

## Label the names of the independent and dependent variables for convenience
x <- c("month","lag12","trend","trend_sqr")
y <- "y"
```

Now that the data has been loaded into the working cluster, we can begin the training process.


### Forecasting with the Random Forest model

Build a forecasting model with the Random Forest (RF) algorithm.

```{r randomForest, echo=FALSE, message=FALSE, warning=FALSE}
## Random Forest model with default settings
bf_rf1 <- h2o.randomForest(training_frame = train_h, # training set
                          nfolds = 5, # number of folders for CV training
                          x=x, # character vector for independent variable names
                          y=y, # string for name of dependent variable
                          ntrees = 500, # number of trees
                          # number of rounds to use before stopping training
                          stopping_rounds = 10,
                          # determine when model should stop and build new trees
                          stopping_metric = "RMSE", # error metric
                          # score during each iteration of model training
                          score_each_iteration = TRUE,
                          # minimal improvement to continue training process
                          stopping_tolerance = 0.0001,
                          seed = 1234)

## View contribution of model inputs
bf_rf1@model$model_summary # review parameters of model performance
h2o.varimp_plot(bf_rf1) # each variable ranked on a scale from 0 to 1

## Plot the learning process of the model as a function of the number of trees
tree_score <- bf_rf1@model$scoring_history$training_rmse
plot_ly(x=seq_along(tree_score),y=tree_score,
        type = "scatter",mode="line")%>%
  layout(title="The Trained Model Score History",
         yaxis=list(title="RMSE"),
         xaxis=list(title="Number of Trees"))

## Measure the model's performance on the testing partition
test_h$pred_rf <- h2o.predict(bf_rf1,test_h) # predict corresponding values
test_1 <- as.data.frame(test_h) # transfer to a data frame object
# calculate MAPE score of RF model on test partition
mape_rf1 <- mean(abs(test_1$y-test_1$pred_rf)/test_1$y)
mape_rf1 # output MAPE score

## Visualize results and compare prediction to actual and baseline predictions
plot_ly(data = test_1)%>% # use test set data frame as input
  add_lines(x=~date,y=~y,name="Actual")%>%
  add_lines(x=~date,y=~yhat,name="Linear Regression",line=list(dash="dot"))%>%
  add_lines(x=~date,y=~pred_rf,name="Random Forest",
            line=list(dash="dash"))%>%
  layout(title=
           "Average Beef Price - Actual vs. Prediction (Random Forest)",
         yaxis=list(title="Cost per Pound"),
         xaxis=list(title="Month"))
```

The Random Forest model with its default settings has an MAPE rate of 11.48%, which is higher than the 10.04% error rate of the benchmark model.


### Forecasting with the GBM model

```{r gradientBoostingMachine, echo=FALSE, message=FALSE, warning=FALSE}
## Train the GBM model with the same input data used previously
bf_gb1 <- h2o.gbm(training_frame = train_h, # training set
                  nfolds = 5, # number of folders for CV training
                  x=x, # character vector for independent variable names
                  y=y, # string for name of dependent variable,
                  max_depth = 20, # set the maximum tree depth
                  distribution = "gaussian",
                  ntrees = 500, # number of trees
                  learn_rate = 0.1, # default value between 0 and 1
                  # score the model during each iteration of training
                  score_each_iteration = TRUE)

## Review the rank of the importance of the model's variables
bf_rf1@model$model_summary # review parameters of model performance
h2o.varimp_plot(bf_gb1) # each variable ranked on a scale from 0 to 1

## Plot the learning process of the model as a function of the number of trees
tree_score <- bf_gb1@model$scoring_history$training_rmse
plot_ly(x=seq_along(tree_score),y=tree_score,
        type = "scatter",mode="line")%>%
  layout(title="The Trained Model Score History",
         yaxis=list(title="RMSE"),
         xaxis=list(title="Number of Trees"))

## Test the model's performance on the testing set
test_h$pred_gbm <- h2o.predict(bf_gb1,test_h) # predict corresponding values
test_1 <- as.data.frame(test_h) # transfer to a data frame object
# calculate MAPE score of GBM model on test partition
mape_gb1 <- mean(abs(test_1$y-test_1$pred_gbm)/test_1$y)
mape_gb1 # output MAPE score

## Visualize results and compare prediction to actual and baseline predictions
plot_ly(data = test_1)%>% # use test set data frame as input
  add_lines(x=~date,y=~y,name="Actual")%>%
  add_lines(x=~date,y=~yhat,name="Linear Regression",line=list(dash="dot"))%>%
  add_lines(x=~date,y=~pred_gbm,name="Gradient Boosting Machine",
            line=list(dash="dash"))%>%
  layout(title=
           "Average Beef Price - Actual vs. Prediction (Gradient Boosting Machine)",
         yaxis=list(title="Cost per Pound"),
         xaxis=list(title="Month"))
```

The Gradient Boosting Machine model with its default settings has an MAPE rate of 10.18%, which is higher than the 10.04% error rate of the benchmark model.


### Prediction for March 1, 2023

```{r predictions}
## Linear regression models
data.frame(forecast(bf_lr1,h=24))[24,] # 4.819455

## Exponential smoothing models
data.frame(forecast(bf_hw1,h=24))[24,] # 4.228599
data.frame(forecast(bf_hw2,h=24))[24,] # 3.827890

## ARIMA models
data.frame(forecast(arima_m1,h=12))[12,] # 1.557115
data.frame(forecast(arima_m2,h=12))[12,] # 1.556631
data.frame(forecast(arima_m3,h=12))[12,] # 1.556562
data.frame(forecast(arima_a1,h=24))[24,] # 4.190090
data.frame(forecast(arima_a2,h=24))[24,] # 4.211082
```
